Simple repression draft

### 2020-09-24 doc/resubmission/appendix_01_nonbursty.tex
update resubmission version of paper

In this section we detail the calculation of mean mRNA levels, fold-changes in
expression, and Fano factors for nonequilibrium promoter models 1 through 4
in Figure~\ref{fig1:means_cartoons}. These are the results that were quoted but
not derived in Sections~\ref{section_02_means} and \ref{sec:beyond_means} of the
main text. In each of these four models, the natural mathematicization of their
cartoons is as a chemical master equation such as Eq.~\ref{eq:2state_rep_cme}
for model 1. Before jumping into the derivations of the general computation of
the mean mRNA level and the Fano factor we will work through the derivation of
an example master equation. In particular we will focus on model 1 from
Figure~\ref{fig1:means_cartoons}(C). The general steps are applicable to all
other chemical master equations in this work.

\mrm{THIS NEEDS TO BE FIXED}
\subsubsection{The Two-State Equilibrium Model}
In this simplest model, depicted as (1) in Figure~\ref{fig1:means_cartoons}(B),
the promoter is idealized as existing in one of two states, either repressor
bound or repressor unbound. The rate of transcription
is assumed to be proportional to the fraction of time spent
in the repressor unbound state.
From the relative statistical weights listed in
Figure~\ref{fig1:means_cartoons}, the probability $p_U$ of being in the
unbound state is
\begin{equation}
p_U = \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1}.
\end{equation}
The mean rate of transcription is then given by $r p_U$, as assumed by
Eq.~\ref{eq:transcrip_prop_pbound}. The mean number of mRNA is
set by the balance of average mRNA transcription and degradation rates,
so it follows that the mean mRNA level is given by
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1},
\end{equation}
where $r$ is the transcription rate from the repressor unbound state, $\gamma$
is the mRNA degradation rate, $R$ is repressor copy number, $N_{NS}$ is the
number of nonspecific binding sites in the genome where repressors spend most
of their time when not bound to the operator, $\beta \equiv 1/k_BT$, and
$\Delta\varepsilon_R$ is the binding energy of a repressor to its operator site.
The derivation of this result is deferred to Appendix~\ref{sec:non_bursty}.

@@

\mrm{THIS NEEDS TO BE FIXED}
\subsubsection{The Poisson Promoter Nonequilibrium Model}
For our first kinetic model, we  imitate the states considered in the Two-State
Equilibrium Model and consider the simplest possible picture with only two
states, repressor bound and unbound. This is exactly the model used for the main
results of~\cite{Jones2014}. In this picture, repressor association and
dissociation rates from its operator site, $k_R^+$ and $k_R^-$, respectively,
govern transitions between the two states. When the system is in the unbound
state, transcription initiates at rate $r$, which represents a coarse-graining
of all the downstream processes into a single effective rate. mRNA is degraded
at rate $\gamma$ as already exploited in the previous models.

>>>>

In this section we detail the calculation of mean mRNA levels, fold-changes in
expression, and Fano factors for both thermodynamic and kinetic promoter models
in Figure~\ref{fig1:means_cartoons}. These are the results that were quoted but
not derived in Sections~\ref{section_02_means} and \ref{sec:beyond_means} of the
main text. In each of these models, the natural mathematicization of their
cartoons is either an equilibrium model based on statisticla mechanics, or a
chemical master equation. These derivations will go through the specifics of the
models in Figure~\ref{fig1:means_cartoons}. We point the readers to some other
excellent reviews on the general
frameworks~\cite{Bintu2005a,Sanchez2013,Phillips2015a}.

\mrm{THIS NEEDS TO BE FIXED}
\subsection{Thremodynamic models of gene regulation}
The first class of models we will explore are the so-called thermodynamic models
of gene regulation~\cite{Ackers1982}. The premise for these models is that we
imagine the steps leading to transcription --binding and unbinding of
transcription factors and of RNAP-- to occur in a timescale much faster than the
downstream processes --open complex formation and RNAP
escape~\cite{Browning2004}. What this allows us to do is to assume that the
steps leading to transcription can be modeled as being in quasi-equilibrium.
This is enormously advantageous since we can use the theoretical tools of 
equilibrium statisticla mechanics to model such process.

This means that we can enlist the possible microstates in which we can find the
promoter in order to calculate their probabilities based on quantities such as
molecular counts and interaction energies. We then assume that the gene
expression level is proportional to the probability of finding the promoter in
the transcriptionally active microstate. As mentioned in the main text, these
models can only deal with the mean gene expression level. This is because the
probability space we consider is that of the state of the repressor, rather than
both the state of the repressor and the mRNA counts. Let's now work through
models 1 and 2 from Figure~\ref{fig1:means_cartoons}(B).

\subsubsection{The Two-State Equilibrium Model}
In this simplest model, depicted as (1) in Figure~\ref{fig1:means_cartoons}(B),
the promoter is idealized as existing in one of two states, either repressor
bound or repressor unbound. The rate of transcription
is assumed to be proportional to the fraction of time spent
in the repressor unbound state.
From the relative statistical weights listed in
Figure~\ref{fig1:means_cartoons}, the probability $p_U$ of being in the
unbound state is
\begin{equation}
p_U = \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1}.
\end{equation}
The mean rate of transcription is then given by $r p_U$, as assumed by
Eq.~\ref{eq:transcrip_prop_pbound}. The mean number of mRNA is
set by the balance of average mRNA transcription and degradation rates,
so it follows that the mean mRNA level is given by
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1},
\end{equation}
where $r$ is the transcription rate from the repressor unbound state, $\gamma$
is the mRNA degradation rate, $R$ is repressor copy number, $N_{NS}$ is the
number of nonspecific binding sites in the genome where repressors spend most
of their time when not bound to the operator, $\beta \equiv 1/k_BT$, and
$\Delta\varepsilon_R$ is the binding energy of a repressor to its operator site.
The derivation of this result is deferred to Appendix~\ref{sec:non_bursty}.

@@

\mrm{THIS NEEDS TO BE FIXED}

Before jumping into the derivations of the general
computation of the mean mRNA level and the Fano factor we will work through the
derivation of an example master equation. In particular we will focus on model 1
from Figure~\ref{fig1:means_cartoons}(C). The general steps are applicable to
all other chemical master equations in this work.

\subsubsection{The Poisson Promoter Nonequilibrium Model}
For our first kinetic model, we  imitate the states considered in the Two-State
Equilibrium Model and consider the simplest possible picture with only two
states, repressor bound and unbound. This is exactly the model used for the main
results of~\cite{Jones2014}. In this picture, repressor association and
dissociation rates from its operator site, $k_R^+$ and $k_R^-$, respectively,
govern transitions between the two states. When the system is in the unbound
state, transcription initiates at rate $r$, which represents a coarse-graining
of all the downstream processes into a single effective rate. mRNA is degraded
at rate $\gamma$ as already exploited in the previous models.

### 2020-09-24 doc/resubmission/appendix_03_bayesian_inference.tex
updates before bringing back Dropbox documents
update resubmission version of paper

\begin{figure}[p]
\centering
\includegraphics{../figures/si/figS0X_ppc.pdf}
\caption{\textbf{Theory-data comparison of inference on unregulated promoters.}
Comparison of the inference (red shaded area) vs the experimental measurements
(black lines) for 18 different unregulated promoters with different mean mRNA
expression levels from Ref.~\cite{Jones2014}. Upper panels show the empirical
cumulative distribution function (ECDF), while the lower panels show the
differences with respect to the median of the posterior samples. White numbers
are the same as in Figure~\ref{fig1:means_cartoons} for cross comparison. The
predicted binding energies $\beta\Delta\varepsilon_p$ were obtained from the
energy matrix model in Ref.~\cite{Brewster2012}}
\label{figS:ppc_unreg}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics{../figures/si/figS0X_aTc_circuit.pdf}
\caption{\textbf{aTc controlled expression of LacI repressor.} Schematic of the
circuit used in~\cite{Jones2014} to control the expression of the LacI
repressor. The \textit{lacI} gene is under the control of the TetR repressor. As
the TetR repressor is inactivated upon binding of anhydrotetracycline or aTc,
the more aTc added to the media were cells are growing, the less TetR repressors
available to control the expression of the \textit{lacI} gene, resulting in more
LacI repressors per cell. LacI simultaneously controls the expression of the
mRNA on which single-molecule mRNA FISH was performed for gene expression
quantification.}
\label{figS:aTc_circuit}
\end{figure}

@@

>>>>

\begin{figure}[p]
\centering
\includegraphics{../../figures/si/figS0X_ppc.pdf}
\caption{\textbf{Theory-data comparison of inference on unregulated promoters.}
Comparison of the inference (red shaded area) vs the experimental measurements
(black lines) for 18 different unregulated promoters with different mean mRNA
expression levels from Ref.~\cite{Jones2014}. Upper panels show the empirical
cumulative distribution function (ECDF), while the lower panels show the
differences with respect to the median of the posterior samples. White numbers
are the same as in Figure~\ref{fig1:means_cartoons} for cross comparison. The
predicted binding energies $\beta\Delta\varepsilon_p$ were obtained from the
energy matrix model in Ref.~\cite{Brewster2012}}
\label{figS:ppc_unreg}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics{../../figures/si/figS0X_aTc_circuit.pdf}
\caption{\textbf{aTc controlled expression of LacI repressor.} Schematic of the
circuit used in~\cite{Jones2014} to control the expression of the LacI
repressor. The \textit{lacI} gene is under the control of the TetR repressor. As
the TetR repressor is inactivated upon binding of anhydrotetracycline or aTc,
the more aTc added to the media were cells are growing, the less TetR repressors
available to control the expression of the \textit{lacI} gene, resulting in more
LacI repressors per cell. LacI simultaneously controls the expression of the
mRNA on which single-molecule mRNA FISH was performed for gene expression
quantification.}
\label{figS:aTc_circuit}
\end{figure}

@@

As suggested by one of our reviewers, in order for readers to judge the
agreement between our predictions and the experimental data for the regulated
case, we include Figure~\ref{figS:eg_histograms} that plot the same data as
Figure~\ref{fig4:repressed_post_full}(C), but rather than showing ECDF, we show
histograms of the individual distributions.

\begin{figure}[p]
\centering
\includegraphics[width=\textwidth]{../../figures/si/figS0X_histograms.pdf}
\caption{\textbf{Theory-experiment comparision of mRNA distributions for
regulated promoters.} Comparison of the inference (color shaded area
encloses the 95\% of all posterior predictive check samples while the color
solid line represents the median of all samples) vs the experimental
measurements (black lines) for different regulated promoters with different
operators (columns) and aTc concentrations (rows) from
Ref.~\cite{Jones2014}.}
\label{figS:reg_histograms}
\end{figure}

### 2020-09-24 doc/resubmission/section_01_introduction.tex
implement Rob's comments
update resubmission version of paper

Gene expression presides over much of the most important dynamism of living
organisms. The level of expression of batteries of different genes is altered as
a result of spatiotemporal cues that integrate chemical, mechanical and other
types of signals. As our ability to experimentally observe and measure the
dynamical processes that constitute the central dogma improves, there is an
opportunity to undertake a theory-experiment dialogue in order to sharpen our
understanding of such a fundamental biological process. One of the remaining
outstanding challenges to have emerged in the genomic era is our continued
inability to predict the regulatory consequences of different regulatory
architectures, i.e. the arrangement and affinity of binding sites for
transcription factors and RNA polymerases on the DNA. This challenge stems first
and foremost from our ignorance about what those architectures even are, with
more than 60\% of the genes even in an ostensibly well understood organism such
as {\it E. coli} having no regulatory insights at all~\cite{Rydenfelt2014-2,
Belliveau2018, Ghatak2019, Santos_Zavaleta2019}. But even once we have
established the identity of key transcription factors and their binding sites of
a given promoter architecture, there remains the predictive challenge of
understanding its input-output properties, an objective that can be met by a
myriad of approaches using the tools of statistical physics~\cite{Ackers1982,
Shea1985, Buchler2003, Vilar2003a, Vilar2003b, Bintu2005a, Bintu2005c,
Gertz2009, Sherman2012, Saiz2013, Ko1991, Peccoud1995, Record1996, Kepler2001,
Sanchez2008, Shahrezaei2008, Sanchez2011, Michel2010, Iyer-Biswas2009}. One
route to such predictive understanding is to focus on the simplest regulatory
architecture and to push the theory-experiment dialogue to increase the
predictive power of our theoretical models~\cite{Garcia2011, Phillips2019}. If
we demonstrate that we can pass that test by successfully predicting both the
means and variance in gene expression at the mRNA level, then that provides a
more solid foundation upon which to launch into more complex problems - for
instance, some of the previously unknown architectures uncovered
in~\cite{Belliveau2018} and~\cite{Ireland2020}.

To that end, in this paper we examine a wide variety of distinct models for the
simple repression regulatory architecture. This genetic architecture consists of
a DNA promoter regulated by a transcriptional repressor that binds to a single
binding site as developed in pioneering early work on the quantitative
dissection of transcription \cite{Oehler1994, Oehler1990}. One of the main
features of of the models we explore is that, by construction, all features
related to the microstates in which the repressor is bound to the promoter can
be separated from the microstates in which the RNA polymerase (RNAP) is bound.
From a modeling perspective, this means that some of the models can be written
as effective two-state models for which there is a rich
literature~\cite{Peccoud1995,Bintu2005a,Shahrezaei2008,Iyer-Biswas2009,Sanchez2013,Jones2014}.
Here, we systematically compare the predictions of several models with different
levels of coarse graining written in terms of thermodynamic and kinetic
parameters. One goal in exploring such coarse-grainings is to build towards the
future models of regulatory response that will be able to serve the powerful
predictive role needed to take synthetic biology from a brilliant exercise in
enlightened empiricism to a rational design framework as in any other branch of
engineering. More precisely, we want phenomenology in the sense of
coarse-graining away atomistic detail, but still retaining biophysical meaning.
In particular a key question is: at this level of coarse-graining, what
microscopic details do we need to explicitly model, and how do we figure that
out? For example, do we need to worry about all or even any of the steps that
individual RNA polymerases go through each time they make a transcript? Turning
the question around, can we see any imprint of those processes in the available
data? If the answer is no, then those processes are irrelevant for our purposes.
Forward modeling and inverse (statistical inferential) modeling are necessary to
tackle such questions. We combine both approaches in order to discard models 
that cannot empirically satisfy the main features of experimental data. First we
apply forward modeling to demonstrate that none of the models are 
distinguishable at the level of mean gene expression. We then extend the 
modeling to look at higher moments of the distribution, eliminating models that
do not empirically satisfy the observed cell-to-cell variability. Finally we 
arrive to a minimal model on which we can apply inverse modeling in order to 
infer the parameters that explain the data.

@@

The logic of the remainder of the paper is as follows. In
section~\ref{section_02_means}, we show how both thermodynamic models
(Figure~\ref{fig1:means_cartoons}(B)) and kinetic models based upon the chemical
master equation (Figure~\ref{fig1:means_cartoons}(C)) all culminate in the same
underlying functional form for the fold-change in the average level of gene
expression as shown in Figure~\ref{fig1:means_cartoons}(D).
Section~\ref{sec:beyond_means} goes beyond an analysis of the mean gene
expression by asking how the same models presented in
Figure~\ref{fig1:means_cartoons}(C) can be used to explore noise in gene
expression. To make contact with experiment, all of these models must make a
commitment to some numerical values for the key parameters found in each such
model. Therefore in Section~\ref{section_04_bayesian_inference} we explore the
use of Bayesian inference to establish these parameters and to rigorously answer
the question of how to discriminate between the different models.

>>>>

Gene expression presides over much of the most important dynamism of living
organisms. The level of expression of batteries of different genes is altered as
a result of spatiotemporal cues that integrate chemical, mechanical and other
types of signals. As our ability to experimentally observe and measure the
dynamical processes that constitute the central dogma improves, there is an
opportunity to undertake a theory-experiment dialogue in order to sharpen our
understanding of such a fundamental biological process. One of the remaining
outstanding challenges to have emerged in the genomic era is our continued
inability to predict the regulatory consequences of different regulatory
architectures, i.e. the arrangement and affinity of binding sites for
transcription factors and RNA polymerases on the DNA. This challenge stems first
and foremost from our ignorance about what those architectures even are, with
more than 60\% of the genes even in an ostensibly well understood organism such
as {\it E. coli} having no regulatory insights at all~\cite{Rydenfelt2014-2,
Belliveau2018, Ghatak2019, Santos_Zavaleta2019}. But even once we have
established the identity of key transcription factors and their binding sites of
a given promoter architecture, there remains the predictive challenge of
understanding its input-output properties, an objective that can be met by a
myriad of approaches using the tools of statistical physics~\cite{Ackers1982,
Shea1985, Buchler2003, Vilar2003a, Vilar2003b, Bintu2005a, Bintu2005c,
Kuhlman2007, Gertz2009, Sherman2012, Saiz2013, Ko1991, Peccoud1995, Record1996,
Kepler2001, Sanchez2008, Shahrezaei2008, Sanchez2011, Michel2010,
Iyer-Biswas2009}. One route to such predictive understanding is to focus on the
simplest regulatory architecture and to push the theory-experiment dialogue to
increase the predictive power of our theoretical models~\cite{Garcia2011,
Phillips2019}. If we demonstrate that we can pass that test by successfully
predicting both the means and variance in gene expression at the mRNA level,
then that provides a more solid foundation upon which to launch into more
complex problems - for instance, some of the previously unknown architectures
uncovered in~\cite{Belliveau2018} and~\cite{Ireland2020}.

To that end, in this paper we examine a wide variety of distinct models for the
simple repression regulatory architecture. This genetic architecture consists of
a DNA promoter regulated by a transcriptional repressor that binds to a single
binding site as developed in pioneering early work on the quantitative
dissection of transcription \cite{Oehler1994, Oehler1990}. One of the main
features of of the models we explore is that, by construction, all features
related to the microstates in which the repressor is bound to the promoter can
be separated from the microstates in which the RNA polymerase (RNAP) is bound.
From a modeling perspective, this means that some of the models can be written
as effective two-state models for which there is a rich
literature~\cite{Peccoud1995, Shahrezaei2008, Iyer-Biswas2009, Tkacik2009,
Sanchez2013, Jones2014, So2011, Munsky2012}. Here, we systematically compare the
predictions of several models with different levels of coarse graining written
in terms of thermodynamic and kinetic parameters. One goal in exploring such
coarse-grainings is to build towards the future models of regulatory response
that will be able to serve the powerful predictive role needed to take synthetic
biology from a brilliant exercise in enlightened empiricism to a rational design
framework as in any other branch of engineering. More precisely, we want
phenomenology in the sense of coarse-graining away atomistic detail, but still
retaining biophysical meaning. In particular a key question is: at this level of
coarse-graining, what microscopic details do we need to explicitly model, and
how do we figure that out? For example, do we need to worry about all or even
any of the steps that individual RNA polymerases go through each time they make
a transcript? Turning the question around, can we see any imprint of those
processes in the available data? If the answer is no, then those processes are
irrelevant for our purposes. Forward modeling and inverse (statistical
inferential) modeling are necessary to tackle such questions. We combine both
approaches in order to discard models that cannot empirically satisfy the main
features of experimental data. First we apply forward modeling to demonstrate
that none of the models are distinguishable at the level of mean gene
expression. We then extend the modeling to look at higher moments of the
distribution, eliminating models that do not empirically satisfy the observed
cell-to-cell variability. Finally we arrive at a minimal model on which we can
apply inverse modeling in order to infer the parameters that explain the data.

@@

The logic of the remainder of the paper is as follows. In
section~\ref{section_02_means}, we show how both thermodynamic models
(Figure~\ref{fig1:means_cartoons}(B)) and kinetic models based upon the chemical
master equation (Figure~\ref{fig1:means_cartoons}(C)) all culminate in the same
underlying functional form for the fold-change in the average level of gene
expression with an effective free energy $\Delta F_R$ capturing the regulation
given by the transcription factor, and a term $\rho$ describing the level of
coarse-graining of the transcriptional events as shown in
Figure~\ref{fig1:means_cartoons}(D). Section~\ref{sec:beyond_means} goes beyond
an analysis of the mean gene expression by asking how the same models presented
in Figure~\ref{fig1:means_cartoons}(C) can be used to explore noise in gene
expression. To make contact with experiment, all of these models must make a
commitment to some numerical values for the key parameters found in each such
model. Therefore in Section~\ref{section_04_bayesian_inference} we explore the
use of Bayesian inference to establish these parameters and to rigorously answer
the question of how to discriminate between the different models.

### 2020-09-24 doc/resubmission/section_03_beyond_means.tex
update resubmission version of paper

To develop some intuition, consider model 4 in the limit $k^+ \ll k^- \lesssim
r$, which is roughly satisfied by the parameters inferred
in~\cite{Razo-Mejia2020}. In this limit, the system spends the majority of its
time in the inactive state, occasionally becoming active and making a burst of
transcripts. This should call to mind the well-known phenomenon of
transcriptional bursting, as observed in,
e.g.,~\cite{Golding2005,Chong2014,Sevier2016,Sepulveda2016}. Let us make this
correspondence more precise. The mean dwell time in the active state is $1/k^-$.
While in this state, transcripts are produced at a rate $r$ per unit time. So on
average, $r/k^-$ transcripts are produced before the system switches to the
inactive state. Once in the inactive state, the system dwells there for an
average time $1/k^+$ before returning to the active state and repeating the
process. $r/k^-$ resembles an average burst size, and $1/k^+$ resembles the time
interval between burst events. More precisely, $1/k^+$ is the mean time between
the end of one burst and the start of the next, whereas $1/k^+ + 1/k^-$ would be
the mean interval between the start of two successive burst events, but in the
limit $k^+ \ll k^-$, $1/k^+ + 1/k^- \approx 1/k^+$. Note that this limit ensures
that the waiting time between bursts is approximately exponentially distributed,
with mean set by the only timescale left in the problem, $1/k^+$. If instead it
were the case that $k^+ \sim k^-$, then the waiting time $1/k^+ + 1/k^-$ would
have a peaked distribution, but this does not appear to be the case for any of
datasets from~\cite{Jones2014}.

>>>>

To develop some intuition, consider model 4 in the limit $k^+ \ll k^- \lesssim
r$, which is roughly satisfied by the parameters inferred
in~\cite{Razo-Mejia2020}. In this limit, the system spends the majority of its
time in the inactive state, occasionally becoming active and making a burst of
transcripts. This should call to mind the well-known phenomenon of
transcriptional bursting, as observed in,
e.g.,~\cite{Golding2005,Chong2014,Senecal2014,Sevier2016,Sepulveda2016}. Let us
make this correspondence more precise. The mean dwell time in the active state
is $1/k^-$. While in this state, transcripts are produced at a rate $r$ per unit
time. So on average, $r/k^-$ transcripts are produced before the system switches
to the inactive state. Once in the inactive state, the system dwells there for
an average time $1/k^+$ before returning to the active state and repeating the
process. $r/k^-$ resembles an average burst size, and $1/k^+$ resembles the time
interval between burst events. More precisely, $1/k^+$ is the mean time between
the end of one burst and the start of the next, whereas $1/k^+ + 1/k^-$ would be
the mean interval between the start of two successive burst events, but in the
limit $k^+ \ll k^-$, $1/k^+ + 1/k^- \approx 1/k^+$. Note that this limit ensures
that the waiting time between bursts is approximately exponentially distributed,
with mean set by the only timescale left in the problem, $1/k^+$. If instead it
were the case that $k^+ \sim k^-$, then the waiting time $1/k^+ + 1/k^-$ would
have a peaked distribution, but this does not appear to be the case for any of
datasets from~\cite{Jones2014}.
